---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit attributes file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(dplyr)
library(ggplot2)
```
# dbparser

[![Build Status](https://travis-ci.org/ropensci/dbparser.svg?branch=master)](https://travis-ci.org/ropensci/dbparser) 
[![Build status](https://ci.appveyor.com/api/projects/status/k18sqp55n39f3y5w?svg=true)](https://ci.appveyor.com/project/MohammedFCIS/dbparser)
[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/dbparser)](https://cran.r-project.org/package=dbparser)
[![codecov](https://codecov.io/gh/ropensci/dbparser/branch/master/graph/badge.svg)](https://codecov.io/gh/ropensci/dbparser)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
[![metacran downloads](https://cranlogs.r-pkg.org/badges/grand-total/dbparser)](https://cran.r-project.org/package=dbparser)
[![Rdoc](http://www.rdocumentation.org/badges/version/dbparser)](http://www.rdocumentation.org/packages/dbparser)
[![CII Best Practices](https://bestpractices.coreinfrastructure.org/projects/3311/badge)](https://bestpractices.coreinfrastructure.org/projects/3311)
[![](https://badges.ropensci.org/347_status.svg)](https://github.com/ropensci/software-review/issues/347)

## Introduction

[DrugBank](http://drugbank.ca/) database is a comprehensive, freely
accessible, online database containing information on drugs and drug
targets. As both a bioinformatics and a cheminformatics resource,
DrugBank combines detailed drug (i.e. chemical, pharmacological and
pharmaceutical) data with comprehensive drug target (i.e. sequence,
structure, and pathway) information. More information about DrugBank can
be found [here](https://www.drugbank.ca/about/).

In its raw form, the DrugBank database is a single
XML file. Users must create an [account](https://www.drugbank.ca/public_users/sign_up)
with Drugbank and request permission to [download](https://www.drugbank.ca/releases/latest)
the database. Note that this may take a couple of days.

The `dbparser` package parses the DrugBank XML database into `R` tibbles that can be explored and analyzed by the user, check [this tutorial](https://ropensci.github.io/dbparser/articles/dbparser.html) for more details.

Also, the package offers the option to save these tibbles in databases including **SQL Server DDB** and **Maria DB** just by enabling `save_table` option,  check [this tutorial](https://ropensci.github.io/dbparser/articles/Database_Saving.html) for more details.

If you are waiting for access to the DrugBank database, or do not intend to do a deep dive with
the data, you may wish to use the `dbdataset`
[package](https://ropensci.github.io/dbdataset/index.html), which contains
the DrugBank database already parsed into `R` tibbles. Note that this is a large package that
exceeds the limit set by CRAN. It is only available on GitHub.

`dbparser` is tested against DrugBank versions *5.1.0* through *5.1.4*
successfully. If you find errors with these versions or any other
version please submit an issue
[here](https://github.com/ropensci/dbparser/issues).

## Installation

You can install the released version of dbparser from 
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("dbparser")
```
or you can install the latest updates directly from the repo

``` r
library(devtools)
devtools::install_github("ropensci/dbparser")
```

## Code of Conduct
Please note that the 'dbparser' project is released with a
  [Contributor Code of Conduct](CODE_OF_CONDUCT.md).
  By contributing to this project, you agree to abide by its terms.
  
## Contributing Guide
👍🎉 First off, thanks for taking the time to contribute! 🎉👍
Please review our [Contributing Guide](CONTRIBUTING.md).

## Share the love ❤️

Think **dbparser** is useful? Let others discover it, by telling them in person, via Twitter or a blog post.

Using **dbparser** for a paper you are writing? Consider citing it
```{r}
citation("dbparser")
```
