---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit attributes file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(dplyr)
library(ggplot2)
```
# dbparser

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
[![Build Status](https://travis-ci.org/Dainanahan/dbparser.svg?branch=master)](https://travis-ci.org/Dainanahan/dbparser) 
[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/dbparser)](https://cran.r-project.org/package=dbparser) 
[![DOWNLOADSTOTAL](https://cranlogs.r-pkg.org/badges/grand-total/dbparser)](https://cranlogs.r-pkg.org/badges/grand-total/dbparser)
[![Rdoc](http://www.rdocumentation.org/badges/version/dbparser)](http://www.rdocumentation.org/packages/dbparser)

The main purpose of the `dbparser` package is to parse the 
[DrugBank](http://drugbank.ca/) database which is downloadable in XML format 
from [this link](https://www.drugbank.ca/releases/latest). The parsed data can 
then be explored and analyzed as desired by the user. The `dbparser` package 
further provides the facility of saving the parsed data into a given database.

## Installation

You can install the released version of dbparser from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("dbparser")
```
or you can install the latest updates directly from the repo

``` r
library(devtools)
devtools::install_github("Dainanahan/dbparser")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r}
## parse data from XML and save it to memory
dbparser::get_xml_db_rows(
  "D:/DS Projects/Drug_Interaction/drugbank.xml"
              #system.file("extdata", "drugbank_record.xml", package = "dbparser")
            )

## load drugs data
drugs <- dbparser::parse_drug()

## load drug groups data
drug_groups <- dbparser::parse_drug_groups()

## load drug targets actions data
drug_targets_actions <- dbparser::parse_drug_targets_actions()
```

## Saving into a database

The parsed data may be saved into a given database. Databases supported by 
`dbparser` include MS SQL Server, MySQL and any database supported by `DBI` package. 
Following is an example of saving the 
parsed data into a MySQL database.

```r
library(dbparser)

## open a connection to the desired database engine with an already
## created database
 open_db(xml_db_name =  "drugbank.xml", driver = "SQL Server",
 server = "ServerName\\\\SQL2016", output_database = "drugbank")

## save 'drugs' dataframe to DB
 parse_drug(TRUE)

## save 'drug_groups' dataframe to DB
 parse_drug_groups(TRUE)

## save 'drug_targets_actions' dataframe to DB
 parse_drug_targets_actions(TRUE)

## finally close db connection 
 close_db()
```



## Exploring the data

Following is an example involving a quick look at a few aspects of the parsed 
data. First we look at the proportions of `biotech` and `small-molecule` drugs 
in the data.

```{r}
## view proportions of the different drug types (biotech vs. small molecule)
drugs %>% 
    select(type) %>% 
    ggplot(aes(x = type, fill = type)) + 
    geom_bar() + 
    guides(fill=FALSE)     ## removes legend for the bar colors
```

Below, we view the different `drug_groups` in the data and how prevalent they 
are.

```{r}
## view proportions of the different drug types for each drug group
drugs %>% 
    full_join(drug_groups, by = c('primary_key' = 'drugbank_id')) %>% 
    select(type, group) %>% 
    ggplot(aes(x = group, fill = type)) + 
    geom_bar() + 
    theme(legend.position= 'bottom') + 
    labs(x = 'Drug Group', 
         y = 'Quantity', 
         title="Drug Type Distribution per Drug Group", 
         caption="created by ggplot") + 
    coord_flip()
```

Finally, we look at the `drug_targets_actions` to observe their proportions as 
well.

```{r}
## get counts of the different target actions in the data
targetActionCounts <- 
    drug_targets_actions %>% 
    group_by(action) %>% 
    summarise(count = n()) %>% 
    arrange(desc(count))

## get bar chart of the 10 most occurring target actions in the data
p <- 
    ggplot(targetActionCounts[1:10,], 
           aes(x = reorder(action,count), y = count, fill = letters[1:10])) + 
    geom_bar(stat = 'identity') +
    labs(fill = 'action', 
         x = 'Target Action', 
         y = 'Quantity', 
         title = 'Target Actions Distribution', 
         subtitle = 'Distribution of Target Actions in the Data',
         caption = 'created by ggplot') + 
    guides(fill = FALSE) +    ## removes legend for the bar colors
    coord_flip()              ## switches the X and Y axes

## display plot
p
```
