---
title: "DrugBank Parser"
author: "Mohammed Ali, Ali Ezzat"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DrugBank Parser}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The main purpose of the `dbparser` package is to parse the 
[DrugBank](http://drugbank.ca/) database which is downloadable in XML format 
from [this link](https://www.drugbank.ca/releases/latest). The parsed data can 
then be explored and analyzed as desired by the user. The `dbparser` package 
further provides the facility of saving the parsed data into a given database.

## Getting Started -- Loading the Data

Following is sample code that parses the DrugBank database, then loads the 
drugs info, drug groups info and drug targets actions info.

```{r}
## load dbparser package
library(dbparser)
library(tidyverse)

## parse data from XML and save it to memory
get_xml_db_rows('../full database.xml')

## load drugs data
drugs <- parse_drug()

## load drug groups data
drug_groups <- parse_drug_groups()

## load drug targets actions data
drug_targets_actions <- parse_drug_targets_actions()
```




----------------------------
Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.
----------------------------




## Saving into a database

The parsed data may be saved into a given database. Databases supported by 
`dbparser` include ???, ??? and ???. Following is an example of saving the 
parsed data into a MySQL database.

```{r}
## save 'drugs' dataframe to DB
parse_drug(TRUE)

## save 'drug_groups' dataframe to DB
parse_drug_groups(TRUE)

## save 'drug_targets_actions' dataframe to DB
parse_drug_targets_actions(TRUE)
```




----------------------------
The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css
----------------------------




## Exploring the data

Following is an example involving a quick look at a few aspects of the parsed data.

```{r}
## view proportions of the different drug types (biotech vs. small molecule)
drugs %>% 
    select(type) %>% 
    ggplot(aes(x = type, fill = type)) + 
    geom_bar()
```

```{r}
## view proportions of the different drug types for each drug group
drugs %>% 
    rename(parent_key = primary_key) %>% 
    full_join(drug_groups, by = 'parent_key') %>% 
    select(type, text) %>% 
    ggplot(aes(x = text, fill = type)) + 
    geom_bar() + 
    labs(x = 'Drug Group', 
         y = 'Quantity', 
         title="Drug Type Distribution per Drug Group", 
         caption="created by ggplot") + 
    coord_flip()
```

```{r}
## get counts of the different target actions in the data
targetActionCounts <- 
    drug_targets_actions %>% 
    group_by(text) %>% 
    summarise(count = n()) %>% 
    arrange(desc(count))

## get bar chart of the 10 most occurring target actions in the data
p <- 
    ggplot(targetActionCounts[1:10,], 
           aes(x = reorder(text,count), y = count, fill = letters[1:10])) + 
    geom_bar(stat = "identity") +
    theme(axis.line = element_blank(), 
          plot.title = element_text(hjust=0.5), 
          plot.subtitle = element_text(hjust=0.5)) + 
    labs(fill="action", 
         x = 'Target Action', 
         y = 'Quantity', 
         title="Target Actions Distribution", 
         subtitle = 'Distribution of Target Actions in the Data',
         caption="created by ggplot") + 
    guides(fill=FALSE) +    ## removes legend for the bar colors
    coord_flip()            ## switches the X and Y axes

## display plot
p
```




----------------------------
The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

----------------------------



